---
title: "Bottle data"
author: "Sam Bashevkin"
date: "`r format(Sys.Date(), '%b %d %Y')`"
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
require(leaflet)
require(patchwork)
```

# Introduction

Load data

```{r message=FALSE, warning=FALSE}
d<-read_csv("~/NCEAS training/NPEO2010BottleData.csv")
```

How does salinity vary among stations? Which station is the saltiest and which has the most variable salinity?

# Analysis

Where are these stations?

```{r}
dstations<-d%>%
  select(station, lat, lon)%>%
  distinct()

leaflet(dstations)%>%
  addTiles()%>%
  addCircles(lng=~lon, lat = ~lat)
```


Plot salinity vs depth curves for selected stations.

```{r, echo=FALSE}

  p1<-ggplot(d)+
    geom_point(aes(x=Actual_depth, y=salinity_A, fill=station, shape=station), shape=21)+
    ylab("Salinity")+
    xlab("Depth (m)")+
    theme_bw()+
    theme(panel.grid=element_blank())
  
  p2<-ggplot(d)+
    geom_point(aes(x=Temp_1, y=salinity_A, fill=station, shape=station), shape=21)+
    ylab("Salinity")+
    xlab("Temperature")+
    theme_bw()+
    theme(panel.grid=element_blank())
  
p1|p2
```

Summarise salinity for each station
```{r}
dsal<-d%>%
  group_by(station)%>%
  summarise(Salinity=mean(salinity_A, na.rm=T), Salinity_sd=sd(salinity_A, na.rm=T))

dsal
```

# Results

Salinity ranges from `r min(d$salinity_A, na.rm=T)` to `r max(d$salinity_A, na.rm=T)`. `r dsal$station[which.max(dsal$Salinity)]` is the saltiest station and `r dsal$station[which.min(dsal$Salinity)]` is the freshest station. `r dsal$station[which.max(dsal$Salinity_sd)]` has the most variable salinity and `r dsal$station[which.min(dsal$Salinity_sd)]` has the least variable salinity. 